---
title: "02- tidy w tibbles"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(dplyr)
library(magrittr)
```
# Tibble is the central data structure for the set of packages known as the tidyverse. Tibbles are data frames, but they tweak some older behaviors.
```{r}
vignette("tibble")
```
# Creating a tibble using tribble() function.
# Creating tribble() is a row-wise, readable tibble. This is useful for   creating small tables of data
```{r}
clean_data <- tibble::tribble(~family, ~patient,
  "F1", 101,
  "F1", 102,
  "F2", 201
)

clean_data
```
# The tbl_df (tibble diff) class is a subclass of data.frame, created in order to have different default behavior. The colloquial term "tibble" refers to a data frame that has the tbl_df class. 

```{r}
patient <- tibble(
  subject_id = 1001:1006,
  gender_age = paste(c("m", "f"), 
  floor(runif(6, 21,65)), sep="-"))
  #runif = random uniform values(6 numbers, from=21,to=65)
patient
```

# Key-value pairs: set of two linked data items: a key, which is a unique identifier for some item of data, and the value, which is either the data that is identified or a pointer to the location of that data.
# Convert=TRUE, useful if the component columns are integer, numeric or logical.
```{r}
patient %>% separate(gender_age, c("sex", "age"), convert=TRUE)
```

# Input tibble
```{r}
data_value <- 
   tibble(
   year = c(2015, 2014, 2014),
   month = c(11, 2, 4),
   day = c(23, 1, 30),
   value = c("high", "low", "low")
   ) 

data_value
```
# unite()
# No need to clean up old columns.
# Use package lubridate for working with dates.
```{r}
date_unite <-  unite(data_value, 
           date, year, month, day, 
           sep = "-") 

date_unite
```
# Split with separate()
# Default split on non-alphanumeric characters
# Use quotes to designate columns since we are not referring to objects.

```{r}
date_unite %>%
  separate(date, c("year", "month", "day"))
```

# Multiple values per cell
```{r}
patient_df <- tibble(
  subject_id = 1001:1003, 
  visit_id = c("1,2,3", "1,2", "1"),
  measured = c("9,0,11", "11,3" , "12")
  )
patient_df
```

#Combinations of variables
```{r}
patient_df %>% 
  separate_rows(visit_id, measured, convert=TRUE) -> patient_separate

patient_separate
```

# Combinations of variables

```{r}
patient_separate %>% 
  complete(subject_id, 
           nesting(visit_id)) #nesting - chaining hierarchy, include all possible combinations of values, will fill with NA

#Note - can use 'nesting" with right and anti joins
```

# Determine filling element as list

```{r}
patient_separate %>% 
  complete(subject_id, 
           nesting(visit_id), fill = list(measured = NA)) #can change = value of choice
```
#Use <NA>, not 0 for missing data in real life application.