---
title: "Data Manipulation-GRP Population"
goal: tutorial in dplyr package
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(magrittr)
library(tidyverse)
library(dplyr)
```

```{r data, echo=FALSE}
gapminder <- read.csv("~/Data/GP_population_5Y.csv")
View(gapminder)
```


```{r}
#calculate summary statistics across groups of observations
mean(gapminder[gapminder$continent == "Africa", "gdpPercap"])
mean(gapminder[gapminder$continent == "Americas", "gdpPercap"])
mean(gapminder[gapminder$continent == "Asia", "gdpPercap"])
```

```{r}
#We are interested in a few variables in the dataset. We could use the select() function to keep only the variables we select.

year_country_gdp <- dplyr::select(gapminder,year, country, gdpPercap)

head(year_country_gdp)
```

```{r}
year_country_gdp <- gapminder[,c("year", "country", "gdpPercap")]
head(year_country_gdp)
```

```{r dplyr %>%, warning=TRUE}
year_country_gdp2 <- gapminder %>% dplyr::select(year,country,gdpPercap)

View(year_country_gdp2)
```

## is equivalent to: 
year_country_gdp <- select(gapminder, year, country, gdpPercap)

#We are interested in African countries. We can combine select and filter to select only the observations where "continent" is "Africa".
```{r select and filter}
year_country_gdp_africa <- gapminder %>%
    filter(continent == "Africa") %>%
    dplyr::select(year,country,gdpPercap)
```

#Logicals
#| <  | less than 
#| >  | greater than 
#| == | equal to
#| != | not equal to 
#| <= | less than or equal to 
#| >= | greater than or equal to |
#| %in% | group membership  
#| is.na | is NA             
#| !is.na | is not NA        
#|& &#124; 
#|! xor any all | Boolean Operators 

```{r question 1}
#Create a new dataframe object called "africa_asia_lifeExp" that is filtered by "Africa" and "Asia" and has a lifeExp less than 60*. Select all variables except for "gdpPercap". Use the pipe operator to chain the functions together. 

```

#Calculate the mean GDP per capita for each continent?

#In base R, you would have to run the mean() function for each subset of data:

#mean(gapminder$gdpPercap[gapminder$continent == "Africa"])
#using dplyr::summarize

```{r dplyr/calculations across groups}
gdp_bycontinents <- gapminder %>%
    group_by(continent) %>%
    dplyr::summarize(mean_gdpPercap = mean(gdpPercap))
head(gdp_bycontinents)

```


```{r group_by mult vars}
str(gapminder %>% group_by(continent))

gdp_bycontinents_byyear <- gapminder %>%
    dplyr::group_by(continent, year) %>%
    dplyr::summarize(mean_gdpPercap = mean(gdpPercap))
head(gdp_bycontinents_byyear)
```

group_by() multiple variables and defining multiple variable with summarize()

```{r group_by and summarize}
gdp_pop_bycontinents_byyear <- gapminder %>%
    dplyr::group_by(continent, year) %>%
    dplyr::summarize(mean_gdpPercap = mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop))
head(gdp_pop_bycontinents_byyear)
```

Exercise: Create a data frame containing the median "lifeExp" for each country.
```{r question 2}

```

# dplyr- mutate()

#Adding values to our original data frame instead of creating a new object. 

```{r mutate}
gapminder_with_extra_vars <- gapminder %>%
    dplyr::group_by(continent, year) %>%
    mutate(mean_gdpPercap = mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop))
head(gapminder_with_extra_vars)
```

## dplyr::mutate()
#Use mutate() to create new variables prior to (or even after) summarizing information.

```{r mutate new vars}
gdp_pop_bycontinents_byyear <- gapminder %>%
    mutate(gdp_billion = gdpPercap*pop/10^9) %>% dplyr::group_by(continent, year) %>%
    dplyr::summarize(mean_gdpPercap =  mean(gdpPercap), sd_gdpPercap = sd(gdpPercap),
mean_pop = mean(pop), sd_pop = sd(pop),
mean_gdp_billion = mean(gdp_billion),
sd_gdp_billion = sd(gdp_billion))

head(gdp_pop_bycontinents_byyear)
```

## dplyr - arrange()

```{r arrange}
gapminder_with_extra_vars <- gapminder %>%
    group_by(continent, year) %>%
    mutate(mean_gdpPercap = mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop)) %>%
    arrange(desc(year), continent)
head(gapminder_with_extra_vars)
```

Exercise:
Use dplyr to add a column to the gapminder dataset that contains the total population of the continent of each observation in a given year. For example, if the first observation is Afghanistan in 1952, the new column would contain the population of Asia in 1952. 

```{r}
```

Exercise: Use dplyr to: (a) add a column called `gdpPercap_diff` that contains the difference between the observation's `gdpPercap` and the mean `gdpPercap` of the continent in that year, (b) arrange the dataframe by the column you just created, in descending order (so that the relatively richest country/years are listed first)

```{r}
```

In the gapminder_wide dataset, we have 3 "ID variables" (`continent`, `country`, `year`) and 3 "Observation variables" (`pop`, `lifeExp`, `gdpPercap`)

library(tidyr)
```{r}
gap_wide <- read.csv("~/Data/GP_population_wide.csv")
head(gap_wide)
```

## pivot_longer()
```{r pivot_longer()}
gap_long <- gap_wide %>% 
  pivot_longer(
    cols = 3:38, 
    names_to = "obstype_year",
    values_to = "obs_values"
  )
head(gap_long)
```

We put 3 arguments into the `pivot_longer()` function: 

1. The columns of the old observation variables (here using the column indices 3:38, signalling columns 3 through 38) that we want to gather into one variable. Notice that we don't want to melt down columns 1 and 2, as these are considered "ID" variables. 

2. the name of the new column for the new ID variable (`obstype_year`)

3. the name for the new amalgamated observation variable (`obs_value`) 

## pivot_longer()

We can also select observation variables using:

* variable indices
* variable names
* `a:c` to select all variables between a and c
* `-c` to *exclude* c
* `starts_with(x, ignore.case = TRUE)`: all names that starts with `x`
* `ends_with(x, ignore.case = TRUE)`: all names that ends with `x`
* `contains(x, ignore.case = TRUE)`: all names that contain `x`

```{r starts_with}
# with the starts_with() function
gap_long <- gap_wide %>%
    pivot_longer(
    cols = c(starts_with('gdpPercap'), 
             starts_with('lifeExp'),
             starts_with('pop')),
    names_to = "obstype_year",
    values_to = "obs_values"
  )
head(gap_long)
```
```{r remove column}
gap_long <- gap_wide %>% 
    pivot_longer(
    cols = c(-continent, -country),
    names_to = "obstype_year",
    values_to = "obs_values"
  )
head(gap_long)
```

In our long dataset, "obstype_year" actually contains 2 pieces of information, the observation type ("pop", "lifeExp", "gdpPercap") and the "year".

We can use the separate() function to split the character strings into multiple variables:

```{r}
gap_long_sep <- gap_long %>% 
  separate(obstype_year, into = c('obs_type','year'), sep = "_") %>% 
  mutate(year = as.integer(year))
head(gap_long_sep)
```

The opposite of pivot_longer() is pivot_wider(). It spreads our observation variables back out to make a wider table. We can use this function to spread our gap_long() to the original "medium" format.

```{r pivot_wider}
gap_medium <- gap_long_sep %>% 
  pivot_wider(
    names_from = "obs_type", 
    values_from = "obs_values")
head(gap_medium)
```

We can do a few quick fixes to make this dataset identical to the original "GP_population_5Y" dataset:

```{r}
gapminder <- read.csv("~/Data/GP_population_5Y.csv")
head(gapminder)
```

```{r}
# rearrange columns
gap_medium <- gap_medium %>% dplyr::select(country, year, pop, continent, lifeExp, gdpPercap)
head(gap_medium)

# arrange by country, continent, and year
gap_medium <- gap_medium %>% 
  arrange(country,continent,year)
head(gap_medium)
```

#Exercise:Subset the results from the last exercise to select only the `country`, `year`, and `gdpPercap_diff` columns. Use `tidyr` put it in wide format so that countries are rows and years are columns. You might need to use `ungroup()` before the `select()` function.

```{r}

```

#Exercise: Now turn the dataframe above back into the long format with three columns: `country`, `year`, and `gdpPercap_diff`.
```{r}

```

```{r count}
gapminder %>%
    filter(year == 2002) %>%
    dplyr::count(continent, sort=TRUE)

```

#calculations: the n() function is useful if we wanted to get the standard error of the life expectency per continent.
```{r no}
gapminder %>%
    group_by(continent) %>%
    dplyr::summarize(se_pop = sd(lifeExp)/sqrt(n()))

```

```{r summary operations}
gapminder %>%
    group_by(continent) %>%
    dplyr::summarize(
      mean_le = mean(lifeExp),
      min_le = min(lifeExp),
      max_le = max(lifeExp),
      se_le = sd(lifeExp)/sqrt(n()))
```



```{r mutate if-else}
## keeping all data but "filtering" after a certain condition
# calculate GDP only for people with a life expectation above 25
gdp_pop_bycontinents_byyear_above25 <- gapminder %>% 
    mutate(gdp_billion = ifelse(lifeExp > 25,     gdpPercap * pop / 10^9, NA)) %>%
    group_by(continent, year) %>%
    dplyr::summarize(mean_gdpPercap =  mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop),
              mean_gdp_billion = mean(gdp_billion),
              sd_gdp_billion = sd(gdp_billion))

#summarise() has grouped output by continent. You can override using the ".groups" argument.

#Updating analysis only if certain condition is fulfilled

#For life expectations above 40 years, the gpd to be expected in the future is scaled

gdp_future_bycontinents_byyear_high_lifeExp <- gapminder %>% mutate(gdp_futureExpectation =     ifelse(lifeExp > 40, gdpPercap * 1.5,  gdpPercap)) %>%
    group_by(continent, year) %>%
    dlpyr::summarize(mean_gdpPercap =mean(gdpPercap),
              mean_gdpPercap_expected =       mean(gdp_futureExpectation))

```

```{r}
# Get the start letter of each country
starts.with <- substr(gapminder$country, start = 1, stop = 1)
# Filter countries that start with "A" or "Z"
az.countries <- gapminder[starts.with %in% c("A", "Z"), ]
# Make the plot
ggplot(data = az.countries, aes(x = year, y = lifeExp, color = continent)) +
  geom_line() + facet_wrap( ~ country)
```

```{r}
gapminder %>%
   # Get the start letter of each country
   mutate(startsWith = substr(country, start = 1, stop = 1)) %>%
   # Filter countries that start with "A" or "Z"
   filter(startsWith %in% c("A", "Z")) %>%
   # Make the plot
   ggplot(aes(x = year, y = lifeExp, color = continent)) +
   geom_line() +
   facet_wrap( ~ country)

```
```{r}
gapminder %>%
    # Filter countries that start with "A" or "Z"
	filter(substr(country, start = 1, stop = 1) %in% c("A", "Z")) %>%
	# Make the plot
	ggplot(aes(x = year, y = lifeExp, color = continent)) +
	geom_line() +
	facet_wrap( ~ country)

```

