---
title: "Example 4 Data Manipulation-GRP Population"
goal: tutorial in dplyr package
output: html_document
---
#Knitr, by default, looks in the same directory as your .Rmd file to find any files you need to draw in, like data sets or script files. If you keep your data files and scripts in separate folders (and you should) you need to configure knitr.

#You can also choose the working directory from the menu Tools -> Global Options -> R Markdown. And change the "evaluate chunks in directory" to Current or the same as the directory your Rproj is in.
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
knitr::opts_knit$set(root.dir = "~/Data/")
getwd() #directory where source code is located
```

```{r packages}
library(magrittr)
library(tidyverse)
library(dplyr)
library(tidyr)
```

```{r data}
gapminder <- read.csv("GP_population_5Y.csv")
View(gapminder)
```


```{r mean}
#calculate summary statistics across groups of observations
mean(gapminder[gapminder$continent == "Africa", "gdpPercap"])
mean(gapminder[gapminder$continent == "Americas", "gdpPercap"])
mean(gapminder[gapminder$continent == "Asia", "gdpPercap"])
```

```{r select}
#We are interested in a few variables in the dataset. We could use the select() function to keep only the variables we select.

year_country_gdp <- dplyr::select(gapminder,year, country, gdpPercap)

head(year_country_gdp)
```

```{r using [] syntax}
year_country_gdp <- gapminder[,c("year", "country", "gdpPercap")]
head(year_country_gdp)
```

```{r dplyr select %>%}
year_country_gdp2 <- gapminder %>% dplyr::select(year,country,gdpPercap)

View(year_country_gdp2)
```

## is equivalent to: 
year_country_gdp <- select(gapminder, year, country, gdpPercap)

#We are interested in African countries. We can combine select and filter to select only the observations where "continent" is "Africa".
```{r select and filter}
year_country_gdp_africa <- gapminder %>%
    filter(continent == "Africa") %>%
    dplyr::select(year,country,gdpPercap)
```

#Logicals
#| <  | less than 
#| >  | greater than 
#| == | equal to
#| != | not equal to 
#| <= | less than or equal to 
#| >= | greater than or equal to |
#| %in% | group membership  
#| is.na | is NA             
#| !is.na | is not NA        
#|& &#124; 
#|! xor any all | Boolean Operators 

Exercise: 
Create a new dataframe object called "africa_asia_lifeExp" that is filtered by "Africa" and "Asia" and has a lifeExp less than 60*. Select all variables except for "gdpPercap". Use the pipe operator to chain the functions together.
```{r question 1}
 

```

#Calculate the mean GDP per capita for each continent?

#In base R, you would have to run the mean() function for each subset of data:

#mean(gapminder$gdpPercap[gapminder$continent == "Africa"])
#using dplyr::summarize

```{r dplyr/calculations across groups}
gdp_bycontinents <- gapminder %>%
    group_by(continent) %>%
    dplyr::summarize(mean_gdpPercap = mean(gdpPercap))
head(gdp_bycontinents)

```


```{r group_by mult vars}
str(gapminder %>% group_by(continent))

gdp_bycontinents_byyear <- gapminder %>%
    dplyr::group_by(continent, year) %>%
    dplyr::summarize(mean_gdpPercap = mean(gdpPercap))
head(gdp_bycontinents_byyear)
```

group_by() multiple variables and defining multiple variable with summarize()

```{r group_by and summarize}
gdp_pop_bycontinents_byyear <- gapminder %>%
    dplyr::group_by(continent, year) %>%
    dplyr::summarize(mean_gdpPercap = mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop))
head(gdp_pop_bycontinents_byyear)
```

Exercise: 
Create a data frame containing the median "lifeExp" for each country.
```{r question 2}

```

#Adding values to our original data frame instead of creating a new object. 
## dplyr::mutate()

```{r mutate}
gapminder_with_extra_vars <- gapminder %>%
    dplyr::group_by(continent, year) %>%
    mutate(mean_gdpPercap = mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop))
head(gapminder_with_extra_vars)
```

## dplyr::mutate()
#Use mutate() to create new variables prior to (or even after) summarizing information.

```{r mutate new vars}
gdp_pop_bycontinents_byyear <- gapminder %>%
    mutate(gdp_billion = gdpPercap*pop/10^9) %>% dplyr::group_by(continent, year) %>%
    dplyr::summarize(mean_gdpPercap =  mean(gdpPercap), sd_gdpPercap = sd(gdpPercap),
mean_pop = mean(pop), sd_pop = sd(pop),
mean_gdp_billion = mean(gdp_billion),
sd_gdp_billion = sd(gdp_billion))

head(gdp_pop_bycontinents_byyear)
```

## dplyr::arrange()

```{r arrange}
gapminder_with_extra_vars <- gapminder %>%
    group_by(continent, year) %>%
    mutate(mean_gdpPercap = mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop)) %>%
    arrange(desc(year), continent)
head(gapminder_with_extra_vars)
```

Exercise:
Use dplyr to add a column to the gapminder dataset that contains the total population of the continent of each observation in a given year. For example, if the first observation is Afghanistan in 1952, the new column would contain the population of Asia in 1952. 

```{r question 3}

```

Exercise: 
Use dplyr to: (a) add a column called `gdpPercap_diff` that contains the difference between the observation's `gdpPercap` and the mean `gdpPercap` of the continent in that year, (b) arrange the dataframe by the column you just created, in descending order (so that the relatively richest country/years are listed first)
```{r question 4}

```

In the gapminder_wide dataset, we have 3 "ID variables" (`continent`, `country`, `year`) and 3 "Observation variables" (`pop`, `lifeExp`, `gdpPercap`)

```{r wide ds}
gap_wide <- read.csv("GP_population_wide.csv")
head(gap_wide)
```

## tidyr::pivot_longer()
```{r pivot_longer()}
gap_long <- gap_wide %>% 
  pivot_longer(
    cols = 3:38, 
    names_to = "obstype_year",
    values_to = "obs_values"
  )
head(gap_long)
```

We put 3 arguments into the `pivot_longer()` function: 

1. The columns of the old observation variables (here using the column indices 3:38, signalling columns 3 through 38) that we want to gather into one variable. Notice that we don't want to melt down columns 1 and 2, as these are considered "ID" variables. 

2. the name of the new column for the new ID variable (`obstype_year`)

3. the name for the new amalgamated observation variable (`obs_value`) 

We can also select observation variables using:
* variable indices
* variable names
* `a:c` to select all variables between a and c
* `-c` to exclude c
* `starts_with(x, ignore.case = TRUE)`: all names that starts with `x`
* `ends_with(x, ignore.case = TRUE)`: all names that ends with `x`
* `contains(x, ignore.case = TRUE)`: all names that contain `x`

```{r starts_with}
# with the starts_with() function
gap_long <- gap_wide %>%
    pivot_longer(
    cols = c(starts_with('gdpPercap'), 
             starts_with('lifeExp'),
             starts_with('pop')),
    names_to = "obstype_year",
    values_to = "obs_values"
  )
head(gap_long)
```

```{r remove column}
gap_long <- gap_wide %>% 
    pivot_longer(
    cols = c(-continent, -country),
    names_to = "obstype_year",
    values_to = "obs_values"
  )
head(gap_long)
```

In our long dataset, "obstype_year" actually contains 2 pieces of information, the observation type ("pop", "lifeExp", "gdpPercap") and the "year".

We can use the separate() function to split the character strings into multiple variables:

```{r long}
gap_long_sep <- gap_long %>% 
  separate(obstype_year, into = c('obs_type','year'), sep = "_") %>% 
  mutate(year = as.integer(year))
head(gap_long_sep)
```

The opposite of pivot_longer() is pivot_wider(). It spreads our observation variables back out to make a wider table. We can use this function to spread our gap_long() to the original "medium" format.

```{r pivot_wider}
gap_medium <- gap_long_sep %>% 
  pivot_wider(
    names_from = "obs_type", 
    values_from = "obs_values")
head(gap_medium)
```

A few modifications to make this dataset identical to the original "GP_population_5Y" dataset:

```{r original file}
gapminder <- read.csv("GP_population_5Y.csv")
head(gapminder)
```

```{r rearrange}
# rearrange columns
gap_medium <- gap_medium %>% dplyr::select(country, year, pop, continent, lifeExp, gdpPercap)
head(gap_medium)

# arrange by country, continent, and year
gap_medium <- gap_medium %>% 
  arrange(country,continent,year)
head(gap_medium)
```

Exercise:
Subset the results from the last exercise to select only the `country`, `year`, and `gdpPercap_diff` columns. Use `tidyr` put it in wide format so that countries are rows and years are columns. You might need to use `ungroup()` before the `select()` function.
```{r question 5}

```

Exercise: 
Now turn the dataframe above back into the long format with three columns: `country`, `year`, and `gdpPercap_diff`.
```{r question 6}

```

```{r count}
gapminder %>%
    filter(year == 2002) %>%
    dplyr::count(continent, sort=TRUE)

```

#calculations: the n() function is useful if we wanted to get the standard error of the life expectency per continent.
```{r no}
gapminder %>%
    group_by(continent) %>%
    dplyr::summarize(se_pop = sd(lifeExp)/sqrt(n()))

```

```{r summary operations}
gapminder %>%
    group_by(continent) %>%
    dplyr::summarize(
      mean_le = mean(lifeExp),
      min_le = min(lifeExp),
      max_le = max(lifeExp),
      se_le = sd(lifeExp)/sqrt(n()))
```

```{r mutate if-else}
## keeping all data but "filtering" after a certain condition
# calculate GDP only for people with a life expectation above 25
gdp_pop_bycontinents_byyear_above25 <- gapminder %>% 
    mutate(gdp_billion = ifelse(lifeExp > 25,     gdpPercap * pop / 10^9, NA)) %>%
    group_by(continent, year) %>%
    dplyr::summarize(mean_gdpPercap =  mean(gdpPercap),
              sd_gdpPercap = sd(gdpPercap),
              mean_pop = mean(pop),
              sd_pop = sd(pop),
              mean_gdp_billion = mean(gdp_billion),
              sd_gdp_billion = sd(gdp_billion))
```

#summarise() has grouped output by continent. You can override using the ".groups" argument.
#Updating analysis only if certain condition is fulfilled

```{r ifelse}
#For life expectations above 40 years, the gpd to be expected in the future is scaled

gdp_future_bycontinents_byyear_high_lifeExp <- gapminder %>% mutate(gdp_futureExpectation =     ifelse(lifeExp > 40, gdpPercap * 1.5,  gdpPercap)) %>%
    group_by(continent, year) %>%
    dplyr::summarize(mean_gdpPercap =mean(gdpPercap),
              mean_gdpPercap_expected =       mean(gdp_futureExpectation))

```

```{r starts.with}
# Get the start letter of each country
starts.with <- substr(gapminder$country, start = 1, stop = 1)
# Filter countries that start with "A" or "Z"
az.countries <- gapminder[starts.with %in% c("A", "Z"), ]
# Make the plot
ggplot(data = az.countries, aes(x = year, y = lifeExp, color = continent)) +
  geom_line() + facet_wrap( ~ country)
```

#Looking forward to EDA III: ggplot2
```{r gpplot2}
gapminder %>%
   # Get the start letter of each country
   mutate(startsWith = substr(country, start = 1, stop = 1)) %>%
   # Filter countries that start with "A" or "Z"
   filter(startsWith %in% c("A", "Z")) %>%
   # Make the plot
   ggplot(aes(x = year, y = lifeExp, color = continent)) + #plot
   geom_line() + #geometry
   facet_wrap( ~ country) #facet

gapminder %>%
    # Filter countries that start with "A" or "Z"
	filter(substr(country, start = 1, stop = 1) %in% c("A", "Z")) %>%
	# Make the plot
	ggplot(aes(x = year, y = lifeExp, color = continent)) + #plot
	geom_line() + #geometry
	facet_wrap( ~ country) #facet

```

